---
title: "Survival Analysis with R"
output: html_notebook
---

Load built in data package “diabetic" in the survival package
```{r}
library(survival)

data("diabetic", package = "survival")
head(diabetic)
```

View data dimensions
```{r}
dim(diabetic)
```

******

Kaplan meier estimator
```{r}
km <- survfit(Surv(time, status) ~ 1, data = diabetic) # 1 denotes that one survival curve is created for all 
```

Visualise with ggsurplot() function
```{r warning=FALSE}
library(survminer)

ggsurvplot(km, risk.table = TRUE)
```

Kaplan meier estimator - treatment
```{r}
# compute survival function
km_trt <- survfit(Surv(time, status) ~ trt, data = diabetic)

# visualise with ggsurplot() function
# add a line showing the median survival time
km_plot0 <- ggsurvplot(km_trt, surv.median.line = "hv", title = "With or Without Treatment")
```

Kaplan meier estimator - laser
```{r}
# compute survival function
km_ls <- survfit(Surv(time, status) ~ laser, data = diabetic)

# visualise with ggsurplot() function
# add a line showing the median survival time
km_plot1 <- ggsurvplot(km_ls, surv.median.line = "hv", title = "Type of Laser Treatment")
```

Kaplan meier estimator - age
```{r}
# create bins for age 

# view age distribution
hist(diabetic$age)

# create bins
bin_age <- c(0, 10, 20, 30, 40, 50, 60, Inf)

# name bins
bin_name <- c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60+")

# create a column for age bins
diabetic$age_bin <- cut(diabetic$age, breaks = bin_age, labels = bin_name)

# view data
head(diabetic)
```

```{r}
# compute survival function
km_age <- survfit(Surv(time, status) ~ age_bin, data = diabetic)

# visualise with ggsurplot() function
# add a line showing the median survival time
km_plot2 <- ggsurvplot(km_age, surv.median.line = "hv", title = "Age Group")
```

Kaplan meier estimator - eye
```{r}
# compute survival function
km_eye <- survfit(Surv(time, status) ~ eye, data = diabetic)

# visualise with ggsurplot() function
# add a line showing the median survival time
km_plot3 <- ggsurvplot(km_eye, surv.median.line = "hv", title = "Left or Right Eye Damage")
```

Kaplan meier estimator - risk
```{r}
# compute survival function
km_rk <- survfit(Surv(time, status) ~ risk, data = diabetic)

# visualise with ggsurplot() function
# add a line showing the median survival time
km_plot4 <- ggsurvplot(km_rk, surv.median.line = "hv", title = "Risk Group")
```

view plots
```{r}
# call plots
km_plot0
km_plot1
km_plot2
km_plot3
km_plot4
```

******

Weibull model with laser and treatment
```{r}
wb <- survreg(Surv(time, status) ~ risk + trt, data = diabetic)
```

compute coefficients
```{r}
coef(wb)
```

create "imaginary patients"
```{r}
newdat <- expand.grid(
  trt = levels(factor(diabetic$trt)),
  risk = levels(factor(diabetic$risk)))

newdat$risk <- as.numeric(as.character(newdat$risk)) # convert risk to numeric variable

head(newdat)
```

Compute survival curves
```{r}
surv <- seq(.99, .01, by = -.01)
t <- predict(wb, type = "quantile", p = 1-surv, newdata = newdat)
```

Create a dataframe
```{r}
library(reshape2)

# combine the information in newdat with t
surv_wb_wide <- cbind(newdat, t) 

# bring the data.frame to long format
surv_wb <- melt(surv_wb_wide, id.vars = c("trt", "risk"), variable.name = "surv_id", value.name = "time") 

# surv_id to add the correct survival probabilities surv
surv_wb$surv <- surv[as.numeric(surv_wb$surv_id)]

# add columns upper, lower, std.err, and strata to the data.frame
surv_wb[, c("upper", "lower", "std.err", "strata")] <- NA
```

Visualise with ggsurvplot_df() function
```{r}
ggsurvplot_df(surv_wb, surv.geom = geom_line,linetype = "trt", color = "risk", legend.title = NULL)
```

weibull model for log-normal
```{r}
# Log-Normal model
lnmod <- survreg(Surv(time, status) ~ risk + trt, data = diabetic, dist = "lognormal")

# Newdata
newdat_ln <- expand.grid(
  trt = levels(factor(diabetic$trt)),
  risk = levels(factor(diabetic$risk)))

newdat_ln$risk <- as.numeric(as.character(newdat_ln$risk)) # convert risk to numeric variable

# Survival curve from Weibull model and log-normal model
lnt <- predict(lnmod, type = "quantile", p = 1-surv, newdata = newdat_ln)

# combine the information in newdat with lnt
surv_wide <- cbind(newdat_ln, lnt) 

# Melt the data.frame into long format.
surv_long <- melt(surv_wide, id.vars = c("trt", "risk"), variable.name = "surv_id", value.name = "time")

# Add column for the survival probabilities
surv_long$surv <- surv[as.numeric(surv_long$surv_id)]

# Add columns upper, lower, std.err, and strata contianing NA values
surv_long[, c("upper", "lower", "std.err", "strata")] <- NA

# Plot the survival curves
ggsurvplot_df(surv_long, surv.geom = geom_line, 
              linetype = "trt", color = "risk", legend.title = NULL)
```

Weibull model with all factors
```{r}
wb_all <- survreg(Surv(time, status) ~ risk + trt + eye + age + laser, data = diabetic)

# compute coefficients
coef(wb_all)
```


******

cox model with risk and treatment
```{r}
cox <- coxph(Surv(time, status) ~ risk + trt, data = diabetic)
```

compute coefficients
```{r}
coef(cox)
```
create "imaginary patients"
```{r}
newdat_cox <- expand.grid(
  trt = levels(factor(diabetic$trt)),
  risk = levels(factor(diabetic$risk)))

newdat_cox$risk <- as.numeric(as.character(newdat_cox$risk)) # convert risk to numeric variable

rownames(newdat_cox) <- letters[1:length(newdat_cox$trt)] # include an alphabetical row name for use later

head(newdat_cox)
```

Compute survival curves
```{r}
cxsf <- survfit(cox, data = diabetic, newdata = newdat_cox, conf.type = "none")
```

Create a dataframe
```{r}
# compute data.frame needed for plotting
surv_cx <- surv_summary(cxsf)

# get a character vector of alphabetic index we created earlier as id
id <- as.character(surv_cx$strata)

# multiple of the rows in newdat so that it fits with surv_cx dataframe
m_newdat <- newdat_cox[id, ]

# add surv id info to data.frame
surv_cxmod <- cbind(surv_cx, m_newdat)
```

Visualise with ggsurvplot_df() function
```{r}
ggsurvplot_df(surv_cxmod, linetype = "trt", color = "risk", legend.title = NULL, censor = TRUE)
```

cox model with all factors
```{r}
cox_all <- coxph(Surv(time, status) ~ risk + trt + eye + age + laser, data = diabetic)

# compute coefficients
coef(cox_all)
```

